Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2021-11-02 analysis/paper_draft2.Rmd
revisions

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
author:
  - Sachin S. Pandya
#    orcid_id: 0000-0001-7387-1307
  - Xiaomeng Li
  - Timothy E. Moore
#    orcid_id: 0000-0002-9576-0517
#output: distill::distill_article
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    fig_width: 5
    fig_height: 4
header-includes: 
  - \usepackage{amsmath}
fontsize: 12pt
bibliography: references.bib
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to use historical data on an attorney's use of peremptory strikes in past cases to estimate, in a current trial, how likely an attorney has struck prospective jurors because of cognizable-class membership. We also use the _power prior_ to adjust the weight of such historical data in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with actual historical strike data from a convenience sample of cases from one federal district court.

@@

In the US, individuals selected for jury-service appear in court as scheduled and are questioned by the parties' attorneys and the trial judge. During this process, a prospective juror, if not excused by the trial judge for cause, may still be dismissed if a party's attorney uses one of their limited number of peremptory challenges against them. By asserting a peremptory challenge, a party can declare a prospective juror ineligible ("strikes" that juror) for a seat on the jury without the burden of explaining why.

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, gender, or membership in another cognizable class. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias [@LaFave2020, $\S$ 22.3]. For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class membership was a "factor" in that party's use of strikes [Wash. Rule 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1)]. Peremptory-challenge procedure, and thus the task of proving illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

@@

Here, we demonstrate this approach, including the results of simulation studies to test model performance. Then, we present a software prototype that encodes the same approach with actual historical strike data to help attorneys and others detect bias in peremptory-challenge use in real time during jury selection.

@@

We let $b_{hist}$ differ from $b_{curr}$ (our parameter of interest) to simulate noise associated with historical strike data that is incomplete not at random. Alternatively, this difference simulates a real difference in the bias parameter that depends on a feature that the current trial has but that all the past trials in the historical data do not. To illustrate, suppose prosecutor with a bias value conditional on defendant race: $b > 0$ if the criminal defendant is Black, $b = 0$ if not. If our historical strike data contains only in cases with non-Black defendants, accounting for that historical data may lead us to underestimate prosecutor's bias in the current case ($b_{curr}$) with a Black defendant.

@@

In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the coverage rates, i.e., the proportion of times we can reach the correct conclusion about the bias parameter based on those credible intervals. That is, when the true value of $b_{curr}$ is greater than zero, the credible interval lies to the right of zero. When $b_{curr}$ equals to zero, the credible interval includes zero, and when $b_{curr}$ is less than zero, the credible interval lies to the left of zero. Since we use a weakly-informative prior that pulls the estimate of $b_{current}$ towards zero, we focus not how accurately the model can recover the true value of $b_{curr}$, but on how accurately our model can detect that bias.

@@

Figure 1 depicts the coverage rates of $b_{curr}$ with an 80\%, 90\%, and 95\% credible interval, respectively. Here, $b_{hist}$ denotes the bias parameter for the generated historical strike data and $b_{curr}$ denotes the bias parameter for the current data. The values on the right side represent the size of the historical data. The values on the top of the plots represent the power prior weight parameter $a$. For any given combination, a darker color shade denotes a higher coverage rate, and vice versa.

```{r c80}

@@

@@

C80plot <- coverageplot_func(sims, sims$C80, xlabel, ylabel) +
           labs(title = "(a) Coverage Rates - 80% Credible Interval")

C90plot <- coverageplot_func(sims, sims$C90, xlabel, ylabel) +
           labs(title = "(b) Coverage Rates - 90% Credible Interval")

C95plot <- coverageplot_func(sims, sims$C95, xlabel, ylabel) +
           labs(title = "(c) Coverage Rates - 95% Credible Interval")

@@

```{r fig1a}
c80plot
```

```{r fig1b}
c90plot
```

```{r fig1c}
##| fig.cap=fig1caption

c95plot

@@

Figure 1(a) depicts the coverage rate with an 80\% credible interval. As the first column ($a = 0.1$) depicts, when we place low weight on the historical strike data, we mostly rely on current information. When $b_{current}$ is 3, 2 or -2, -3, the coverage rate is high, very close to 1.

However, when $b_{curr}$ is 1, the coverage rate drops below 0.8. If the bias is not strong, our model can detect it with a lower accuracy. However, as we place more weight on the historical strike data (as $a$ increases), when the historical strike data is consistent with current information, the accuracy of detecting the bias is also close to 1 when the bias parameter is 1. When the bias parameter is 0 (no bias), however, the coverage rate is also around 80\%. Thus, if we use the 80\% credible interval, we have about a 20\% chance of erroneously detecting non-zero bias.

Using a wider credible interval lowers this chance of error. For example, with a 90\% credible interval (Figure 1(b)), when $b = 2$, the color is lighter (lower coverage rate) than with a 80\% credible interval; when $b=0$, the color is darker (higher coverage rate). The tradeoff is lower accuracy. With a 90\% credible interval, when the bias parameter is 2, the model is less accurate in detecting bias. When the bias parameter is 3, however, the coverage rate is still high.

Figure 1(c) depicts coverage rates when using a 95\% credible interval. Here, given $b_{curr} = 0$, the coverage rate is very close to 1. This implies that we are very unlikely to infer bias when there is none. When $b_{curr} = 3$, the coverage rate is still high. This implies that we can still accurately detect the bias when it is strong. However, when the bias is moderate ($b_{curr} = \{2, -2\}$), the coverage rate is only high when we have consistent historical information.

@@

```{r simltns}

#| fig.cap="Simulation results for simultaneous strikes example. Columns are values of bias used to simulate data. Blue points denote estimated posterior mean for the bias parameter. Blue lines denote the 95% credible interval. Red line denotes true value of bias parameters used in simulations."

@@

simplot

@@

To test this premise, we conducted the following simulation study. We first generated a single trial in which one attorney used fifteen strikes in an uninterrupted sequence against 32 prospective jurors, 10 of which were member of a cognizable-class (e.g., Black jurors).  Then, we shuffled the order of those strikes to generate 50 trials with the same proportion of struck cognizable-class members but different orders. We then used the model to fit the 50 trials to see if the estimated bias parameters are close to each other. We conducted the simulation study under three scenarios from moderate bias to strong bias ($b=\{1,2,3\}$). Figure 4 depicts the results.

@@

```{r figapp1, out.width='50%', fig.cap = "Density Plots without Historical Strike Data", fig.align='center'}

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
author:
  - Sachin S. Pandya
#    orcid_id: 0000-0001-7387-1307
  - Xiaomeng Li
  - Timothy E. Moore
#    orcid_id: 0000-0002-9576-0517
#output: distill::distill_article
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    fig_width: 5
    fig_height: 5
header-includes: 
  - \usepackage{amsmath}
fontsize: 12pt
bibliography: references.bib
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to use historical data on an attorney's use of peremptory strikes in past cases to estimate, in a current trial, how likely an attorney has struck prospective jurors because of cognizable-class membership. We also use the _power prior_ to adjust the weight of such historical data in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with actual historical strike data from a convenience sample of cases from one federal district court.

@@

In the U.S., individuals selected for jury-service appear in court as scheduled and are questioned by the parties' attorneys and the trial judge. During this process, a prospective juror, if not excused by the trial judge for cause, may still be dismissed if a party's attorney uses one of their limited number of peremptory challenges against them. By asserting a peremptory challenge, a party can declare a prospective juror ineligible ("strikes" that juror) for a seat on the jury without the burden of explaining why.

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, gender, or membership in another cognizable class. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not [@LaFave2020, $\S$ 22.3(d)]. For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class membership was a "factor" in that party's use of strikes [Wash. Rule 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1)]. Peremptory-challenge procedure, and thus the task of proving illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

@@

Here, we demonstrate this approach with simulation studies to test model performance. Then, we present a software prototype that encodes the same approach with actual historical strike data to help attorneys and others detect bias in peremptory-challenge use in real time during jury selection.

@@

We let $b_{hist}$ differ from $b_{curr}$ (our parameter of interest) to simulate noise associated with historical strike data that is incomplete not at random. Alternatively, this difference simulates a real difference in the bias parameter that depends on a feature of the current trial but not of all the past trials in the historical data. To illustrate, suppose prosecutor with a bias value conditional on defendant race: $b > 0$ if the criminal defendant is Black, $b = 0$ if not. If our historical strike data contains only in cases with non-Black defendants, accounting for that historical data may lead us to underestimate prosecutor's bias in the current case ($b_{curr}$) with a Black defendant.

@@

In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the coverage rates, i.e., the proportion of times we can reach the correct conclusion about the bias parameter based on those credible intervals. That is, when the true value of $b_{curr}$ is greater than zero, the credible interval lies to the right of zero; when $b_{curr}$ equals to zero, the credible interval includes zero; and when $b_{curr}$ is less than zero, the credible interval lies to the left of zero. Since we use a weakly-informative prior that pulls the estimate of $b_{current}$ towards zero, we focus not on how accurately the model can recover the true value of $b_{curr}$, but on how accurately our model can detect that bias.

@@

Figures 1-3 depicts the coverage rates of $b_{curr}$ with an 80\%, 90\%, and 95\% credible interval, respectively. Here, $b_{hist}$ denotes the bias parameter for the generated historical strike data and $b_{curr}$ denotes the bias parameter for the current data. The values on the right side represent the size of the historical data. The values on the top of the plots represent the power prior weight parameter $a$. For any given combination, a darker color shade denotes a higher coverage rate, and vice versa.

```{r plotsfig1, eval = FALSE}

@@

sims <- readr::read_csv(load_path, show_col_types = FALSE)
sims$a.f <- as.factor(sims$a)

@@

C80plot <- coverageplot_func(sims, sims$C80, xlabel, ylabel) +
           labs(title = "Coverage Rates - 80% Credible Interval")

C90plot <- coverageplot_func(sims, sims$C90, xlabel, ylabel) +
           labs(title = "Coverage Rates - 90% Credible Interval")

C95plot <- coverageplot_func(sims, sims$C95, xlabel, ylabel) +
           labs(title = "Coverage Rates - 95% Credible Interval")

# output to figures folder

load_path <- here::here("analysis","figures","C80plot.png")
ggsave(load_path, plot = C80plot)
load_path <- here::here("analysis","figures","C90plot.png")
ggsave(load_path, plot = C90plot)
load_path <- here::here("analysis","figures","C95plot.png")
ggsave(load_path, plot = C95plot)

@@

```{r fig1, fig.cap = "Coverage Rates - 80% Credible Interval", fig.align = "center"}

load_path <- here::here("analysis","figures","C80plot.png")
knitr::include_graphics(load_path)

@@

```{r fig2, fig.cap = "Coverage Rates - 90% Credible Interval", fig.align = "center"}

load_path <- here::here("analysis","figures","C90plot.png")
knitr::include_graphics(load_path)

```{r fig3, fig.cap = "Coverage Rates - 95% Credible Interval", fig.align = "center"}

load_path <- here::here("analysis","figures","C95plot.png")
knitr::include_graphics(load_path)

Figure 1 depicts the coverage rate with an 80\% credible interval. As the first column ($a = 0.1$) depicts, when we place low weight on the historical strike data, we mostly rely on current information. When $b_{current}$ is 3, 2 or -2, -3, the coverage rate is high, very close to 1. However, when $b_{curr}$ is 1, the coverage rate drops below 0.8. If the bias is not strong, our model detects it with a lower accuracy.

As we place more weight on the historical strike data (as $a$ increases), when the historical strike data is consistent with current information, the accuracy of detecting the bias is also close to 1 when the bias parameter is 1. When the bias parameter is 0 (no bias), however, the coverage rate is also around 80\%. Thus, if we use the 80\% credible interval, we have about a 20\% chance of erroneously detecting non-zero bias.

Using a wider credible interval lowers this chance of error. For example, with a 90\% credible interval (Figure 2), when $b = 2$, the coverage rate is lower (the color is lighter) than with a 80\% credible interval; when $b=0$, the color is darker (higher coverage rate). The tradeoff is lower accuracy. With a 90\% credible interval, when the bias parameter is 2, the model is less accurate in detecting bias. When the bias parameter is 3, however, the coverage rate is still high.

Figure 3 depicts coverage rates when using a 95\% credible interval. Here, given $b_{curr} = 0$, the coverage rate is very close to 1. This implies that we are very unlikely to infer bias when there is none. When $b_{curr} = 3$, the coverage rate is still high. This implies that we can still accurately detect the bias when it is strong. However, when the bias is moderate ($b_{curr} = \{2, -2\}$), the coverage rate is only high when we have consistent historical information.

@@

```{r simltns, eval = FALSE}

@@

load_path <- here::here("analysis","figures","simplot.png")
ggsave(load_path, plot = simplot)

@@

```{r fig4}

#| fig.cap="Simulation results for simultaneous strikes example"

load_path <- here::here("analysis","figures","simplot.png")
knitr::include_graphics(load_path)

To test this premise, we conducted the following simulation study. We first generated a single trial in which one attorney used fifteen strikes in an uninterrupted sequence against 32 prospective jurors, 10 of which were member of a cognizable-class (e.g., Black jurors).  Then, we shuffled the order of those strikes to generate 50 trials with the same proportion of struck cognizable-class members but different orders. We then used the model to fit the 50 trials to see if the estimated bias parameters are close to each other. We conducted the simulation study under three scenarios from moderate bias to strong bias ($b=\{1,2,3\}$).

Figure 4 depicts the results. The columns are values of bias used to simulate the data. Blue points denote estimated posterior mean for the bias parameter. Blue lines denote the 95% credible interval. Red lines denote true value of bias parameters in the simulations.

@@

```{r figapp1, out.width='50%', fig.align='center'}

#| fig.cap = "Density Plots without Historical Strike Data"
